\hspace{1cm} Do syntactic and semantic similarity lead to interference effects?\newline Evidence from self-paced reading and event-related potentials using German

### 2023-10-14 paper/main.tex
minor edits to gen disc

The absence of (and evidence against) syntactic interference is surprising because the psycholinguistic literature has clearly established that syntactic constraints play a role in sentence processing. For example, \cite{dillon2013} and \cite{Sturt2003} argued that principle A of the binding theory controls the retrieval process in the antecedent-reflexive dependency, effectively eliminating interference from other overlapping features (such as number) of intervening nouns that cannot be an antecedent for the reflexive. For further evidence for this idea from individual differences, see \cite{yadav2021individual}. Even more relevant to the present study is the fact that \cite{mertzen} did find syntactic interference effects.

@@

\section{General discussion}
The present SPR and ERP experiments consistently showed effects of semantic interference and no effects of syntactic interference. This result was not only found in the primary dependent measures of the experiments, respectively, but also in the comprehension accuracy. Given the replicability / reproducibility crises in psycholx \citep[see e.g., ][]{vasishth2018_signficancefilter}, and other (related) sciences \citep[see e.g., ][]{openscience2015_reproducibility}, the cross-methodological consistency is astonishing. It might be due to the very high sample sizes of the present experiments which granted the present study higher power than most comparable psycholinguistic studies. The higher power led to more precise estimates of the investigated effects with less uncertainty than in previous studies. This also partially explains the comparatively smaller effect sizes in the present experiments than have been previously reported.

\subsection{Encoding and retrieval interference}
In the SPR data, we have found that the semantic interference effect originated at the distractor and persisted throughout following regions. Due to this time course, encoding interference was the best explanation for the interference effect in the reading times of the critical region (see discussion of Experiment 1). This raises the question whether the ERP effect was also caused by encoding interference, i.e., whether the ERP effect originated prior to the critical region. To answer this question, it is necessary to look at the ERPs elicited by words earlier in the sentence, although that is unusual in ERP research. Figure \ref{fig:erp_precrit} shows the ERPs for all conditions at the critical word and the two pre-critical adverbs which were identical across conditions. It is apparent from the ERPs of the pre-pre-critical and pre-critical word that the semantic interference effect found at the critical verb was not present at the words preceding it.

@@

Now that we have seen that the words directly preceding the critical word did not show the semantic interference effect, one might ask which brain response was elicited by the distractor itself (see Figure \ref{fig:erp_distractor})? Visual inspection of the ERPs elicited by the distractor shows several differences between conditions, but no semantic interference effect in the N400 spatio-temporal window. We assume that the differences in the ERPs elicited by the distractor in the different conditions are caused by two factors. Firstly, the high and low syntactic interference conditions have different word orders, i.e., the distractors are embedded in different phrase structures (simple noun phrase vs.\ noun phrase including an adjective within a prepositional phrase, see Example \ref{ex:materials}). This is also apparent from the large difference between high and low syntactic interference conditions before distractor onset. Secondly, the distractors were not identical across conditions, therefore the ERPs in Figure \ref{fig:erp_distractor} are elicited by different words in the high vs.\ low semantic interference conditions. In conclusion, the distractors induced different brain responses across conditions, but these did not linger throughout the rest of the sentence (see Figure \ref{fig:erp_precrit}). %Firstly, there is a large difference prior to distractor onset (before time 0 ms) at all electrode sites. Secondly, there is a difference between 300 and 500\,ms most at anterior (left) electrode sites. We assume that both these differences are due to the different word orders in the high vs. low syntactic interference conditions, i.e., the different phrase structures that the distractors are embedded in (simple noun phrase vs. noun phrase including an adjective within a prepositional phrase, see Example \ref{ex:materials}). Thirdly, there is a late positivity (starting at approx. 500\,ms) most pronounced at centro-parietal electrodes for high semantic vs. low semantic interference conditions. This semantic interference effect in the amplitude of a late positivity could be either due to

>>>>

The absence of (and evidence against) syntactic interference is surprising because the psycholinguistic literature has clearly established that syntactic constraints play a role in sentence processing. For example, \cite{dillon2013} and \cite{Sturt2003} argued that principle A of the binding theory controls the retrieval process in the antecedent-reflexive dependency, effectively eliminating or attenuating interference from other overlapping features (such as number marking) of intervening nouns that cannot be an antecedent for the reflexive. For further evidence for this idea from individual differences, see \cite{yadav2021individual}. Even more relevant to the present study is the fact that \cite{mertzen} did find syntactic interference effects.

@@

\section{General discussion}

Considered together, the present SPR and ERP experiments consistently showed effects of semantic interference and no effects of syntactic interference. This pattern was not only found in the primary dependent measures of the experiments, respectively, but also in the comprehension accuracy. 
The modeling reported here suggests that one way to explain the observed pattern in the context of cue-based retrieval accounts is by assuming that the parser intelligently searches for a subject that is within the same clause. A broader implication is that the human sentence comprehension system may be using the hierarchical syntactic structure to selectively target nouns for retrieval; this is an idea that has independent support in the literature \cite[e.g.,][]{Sturt2003,dillon2013}.

%Given the replicability / reproducibility crises in psycholinguistics \citep[see e.g., ][]{vasishth2018_signficancefilter,jaeger_etal_2020,nicenboim,}, and other (related) sciences \citep[see e.g., ][]{openscience2015_reproducibility,nicenboim_etal_2020,nieuwland_etal_2018}, the cross-methodological consistency is aston. It might be due to the very high sample sizes of the present experiments which granted the present study higher power than most comparable psycholinguistic studies. The higher power led to more precise estimates of the investigated effects with less uncertainty than in previous studies. 
%This also partially explains the comparatively smaller effect sizes in the present experiments than have been previously reported.

In the remainder of this section, we discuss several important issues that relate to previous experiments that investigated syntactic and semantic interference.

\subsection{Encoding and retrieval interference}
In the SPR data, we found that the semantic interference effect originated at the distractor and persisted throughout the following regions. Due to this time course, encoding interference \parencite{Yadavetal2022,hammerly2019grammaticality} is the best explanation for the interference effect in the reading times at the critical region. This raises the question whether the ERP effect was also caused by encoding interference, i.e., whether the ERP effect originated prior to the critical region. To answer this question, it is necessary to look at the ERPs elicited by words earlier in the sentence, This is an unusual step in ERP research but makes sense as an exploratory step in the present context. Figure \ref{fig:erp_precrit} shows the ERPs for all conditions at the critical word and the two pre-critical adverbs which were identical across conditions. It is apparent from the ERPs of the pre-pre-critical and pre-critical word that the semantic interference effect found at the critical verb was not present at the words preceding it.

@@

Now that we have seen that the words directly preceding the critical word did not show the semantic interference effect, one might ask which brain response was elicited by the distractor itself (see Figure \ref{fig:erp_distractor})? An exploratory visual inspection of the ERPs elicited by the distractor shows several differences between conditions, but no semantic interference effect in the N400 spatio-temporal window. We believe that the differences in the ERPs elicited by the distractor in the different conditions are caused by two factors. First, the high and low syntactic interference conditions have different word orders, i.e., the distractors are embedded in different phrase structures (simple noun phrase vs.\ noun phrase including an adjective within a prepositional phrase, see Example \ref{ex:materials}). This is also apparent from the large difference between high and low syntactic interference conditions before distractor onset. Secondly, the distractors were not identical across conditions, therefore the ERPs in Figure \ref{fig:erp_distractor} are elicited by different words in the high vs.\ low semantic interference conditions. In conclusion, the distractors induced different brain responses across conditions, but these did not linger throughout the rest of the sentence (see Figure \ref{fig:erp_precrit}). %Firstly, there is a large difference prior to distractor onset (before time 0 ms) at all electrode sites. Secondly, there is a difference between 300 and 500\,ms most at anterior (left) electrode sites. We assume that both these differences are due to the different word orders in the high vs. low syntactic interference conditions, i.e., the different phrase structures that the distractors are embedded in (simple noun phrase vs. noun phrase including an adjective within a prepositional phrase, see Example \ref{ex:materials}). Thirdly, there is a late positivity (starting at approx. 500\,ms) most pronounced at centro-parietal electrodes for high semantic vs. low semantic interference conditions. This semantic interference effect in the amplitude of a late positivity could be either due to
